Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-07-31 tex/draft.tex
add equations

\usepackage[final]{graphicx}
\usepackage[final]{fixme}
\fxsetup{layout=footnote}
\usepackage[left]{lineno}
\usepackage{tikz-feynman}
\usepackage{amsmath,amsthm,amssymb,mathtools,caption,subcaption,mathrsfs,cancel,multirow,enumitem,setspace,sectsty,cite,hyperref,siunitx,cancel}
\usepackage{fullpage}

@@

    Experimentally, the $\Wpm$, $\Z$, quarks, and charged leptons are massive\fxnote{Gloss over neutrino mass?}. Explicit mass terms for the fermions or gauge bosons would violate the chiral and gauge symmetries of the SM Lagrangian, however, so $\SUtwoLUoneY$ must be broken. Specifically, the nonzero vacuum expectation value (VEV) of the Higgs field spontaneously breaks $\SUtwoLUoneY$ to $\UoneQ$. In the process, three of the four degrees of freedom associated with the complex Higgs doublet become the longitudinal polarizations of $\Wpm$ and $\Z$, giving them mass. The remaining degree of freedom is the physical Higgs boson. Expanding the Higgs field about its VEV produces Yukawa interactions such as those shown below. These terms are interpreted as giving the quarks and charged leptons mass, where the mass of each particle is given by the product of the Higgs VEV and the coupling constant.

    \fxnote{Yukawa terms go here}

@@

\section{Supersymmetry}
    \fxnote{seems like I should mention that naturalness favors lighter sparticles}  
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions\fxnote{I need to be better at explaining exactly what this means}. In its simplest form, SUSY results in one new boson for every SM fermion, one new fermion for every SM boson, and one new Higgs doublet.\fxnote{make sure I understand the new Higgs doublet} The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-$\frac{1}{2}$ superpartners to the SM bosons are given the name of their SM counterpart with ``ino" tacked on to the end (e.g. Higgsino or wino).\fxnote{describe charginos, neutralinos?} Collectively the superpartners to the SM particles are known as sparticles.\fxnote{cite martin}

@@

    \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [inline=(d.base), layered layout, horizontal=b to c] {
        a -- [scalar] b,
        b -- [fermion, half left] c,
        c -- [fermion, half left] b,
        c -- [scalar] d,
    };
    \quad + \quad
    \feynmandiagram [baseline={(current bounding box.center)}, layered layout, horizontal=b to c] {
        a -- [scalar] b,
        b -- [charged scalar, half left] c,
        c -- [charged scalar, half left] b,
        c -- [scalar] d,
    };
    \quad = \quad 0 \quad
    \caption{Caption!\fxnote{cite something}}
    \label{exact_susy_correction}
    \end{center} \end{figure}

    Exact SUSY also requires that sparticles have the same mass as their SM counterparts, so the lack of sparticles in collider searches implies that SUSY must be broken. With broken susy, the same diagram as above produces a correction of the form

    \noindent \begin{equation}
    \delta \mHiggs = .
    \label{higgs_susy_correction}
    \end{equation}\fxnote{look up}

@@

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs. Fig.~\ref{rp_and_rpv_diagrams} shows two possible processes, one where R-parity gaurantees a stable LSP and one where RPV produces a long-lived but stable LSP. A similar situation arises in GMSB models\fxnote{should I describe GMSB?} where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs.\fxnote{cite dimopoulos low energy}\fxnote{}

@@

    \subsection{Electromagnetic Calorimeter}
        \fxnote{cite CMS TDR}
        \fxnote{I define EE & EB but never use those definitions}
        After traversing the inner tracker, particles next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses \num{61200} lead tunstate crystals in the barrel and \num{7324} in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (\SI{80}{\percent} of light within \SI{25}{ns}), compact (radiation length = \SI{0.89}{cm}), fine-grained (Moli\`ere radius = \SI{2.2}{cm}), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (\SI{30}{photon\per\mega\electronvolt}), which necesitates photodetectors with intrinsic gain that work in magnetic fields.

@@

    \subsection{Hadronic Calorimeter}
        \fxnote{cite CMS TDR}
        \fxnote{define HB & HF never to be used}
        The hadronic calorimeter (HCAL) uses \SI{3.7}{mm} plates of plastic scintillator interspersed within brass absorber to reconstruct the energy deposited during hadronic showers. Embedded wavelength-shifting fibers capture the scintillation light and transfer it to clear fibers to be read out by hybrid photodiodes.

@@

\section{LHC Searches for Long-lived Particles}
        \fxnote{Include diagrams for signal models} 
        \fxnote{mention which data was used for each search} 
        \fxnote{Steal limit plots and any cool figures from papers} 
    \subsection{Overview}
        \fxnote{steal jamie's cartoon?}  
        LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays, so LLP searches are often forced\fxnote{the searchers are forced, not the searches} to stray from the traditional analysis path in terms of object reconstruction\fxnote{really reconstruction? Maybe object definition?} and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that come to a stop before decaying. See Fig.~\ref{jamie_cartoon} for a cartoon of several possible LLP signatures\fxnote{probably a better way to word this sentence}.

@@

    \subsection{Displaced Vertices}
        \fxnote{cite CONF-2017-026}
        \fxnote{Specify dataset}
        LLPs that decay before reaching the inner tracker can be identified with a displaced vertex (DV)\fxnote{DVs defined clearly enough in overview?}\fxnote{Is it still a DV if it decays in active material somewhere?}. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity.\fxnote{make sure I can explain why met and multiplicity} The search is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range\fxnote{Am I defining sensitivity range consistently across analyses?}.

@@

        After estimating $\num{0.02} \pm \num{0.02}$ background events, exactly zero are observed. The result is then interpreted as an upper limit on the gluino mass and pair-production cross section as a function of gluino lifetime, shown in Fig.~\ref{displaced_limit}. For an LSP neutralino mass of \SI{100}{\giga\electronvolt}, the gluino mass limit is approximately \SI{2000}{\giga\electronvolt} for lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range. \fxnote{Include plots}

@@

    \subsection{Disappearing Tracks}
        \fxnote{cite CONF-2017-017}
        \fxnote{nice disappearing track cartoon}
        \fxnote{nice cartoon feynman diagrams of signal processes}
        \fxnote{Specify dataset}
        Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos\fxnote{have these words been defined?}, such as many AMSB models\fxnote{why amsb? what is amsb?}, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate\fxnote{why? how typical?}) that eventaully decays to the neutralino LSP and, commonly, a pion\fxnote{why?}. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be reconstructed, which results in a disappearing track.

@@

        The authors consider two signal processes, both shown in Fig.\ref{disappearing_processes}: (1) electroweak gaugino production resulting in a disappearing track, an ISR jet, and MET and (2) gluino pair production resulting in a disappearing track, four jets, and MET\fxnote{add diagrams}. Process (1) could be the only gaugino production mode at the LHC if the gluino mass is kinematically unaccessable\fxnote{why?}, but process (2), which leads to chargino production in cascade decays, will be dominant if the gluino mass is within reach.

@@

    \subsection{Stopped Particles}
        \fxnote{Add calorimeter search?}  
        \fxnote{Specify dataset}
        \fxnote{cite CMS PAS EXO-17-004}
        \fxnote{mention how stopped particles complement HSCP search?}
        \fxnote{too redundant with overview?}
        With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states using the full 2015 and 2016 datasets.

@@

    \subsection{Heavy Stable Charged Particles}
        \fxnote{cite CMS EXO-15-010}
        \fxnote{Specify dataset}
        \fxnote{nice dE/dx plot (fig 1)}
        Particles that live for more than a couple ns\fxnote{how do I typeset standalone units?} \fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material\fxnote{redundant w/ overview. Probably better explained here than there}. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss, as shown in Fig.~\ref{hscp_dedx}.

@@

        HSCPs will often be misidentified or missed entirely by standard reconstruction algorithms and analysis criteria, which tend to assume $\beta \approx \num{1}$ and $\lvert q \rvert = e$. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models\fxnote{do I need figures for these?}: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with arbitrary electric charge\fxnote{insert feynman diagrams}. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY\fxnote{why?}.

@@

    \fxnote{legible when printed?}  
    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=0.8\textwidth]{images/atlas_susy_summary}
    \caption{Summary of current experimental limits on SUSY models from ATLAS searches.\fxnote{cite something}}
    \label{atlas_susy_summary}
    \end{center} \end{figure}

>>>>

\usepackage[draft]{graphicx}
\usepackage[final]{fixme}
\fxsetup{layout=footnote}
\usepackage[left]{lineno}
\usepackage{tikz-feynman}
\usepackage{amsmath,amsthm,amssymb,mathtools,caption,subcaption,mathrsfs,cancel,multirow,enumitem,setspace,sectsty,cite,hyperref,siunitx,cancel}
\usepackage{fullpage}

@@

    Experimentally, the $\Wpm$, $\Z$, quarks, and charged leptons are massive\fxnote{Gloss over neutrino mass?}. Explicit mass terms for the fermions or gauge bosons would violate the chiral and gauge symmetries of the SM Lagrangian, however, so $\SUtwoLUoneY$ must be broken. Specifically, the nonzero vacuum expectation value (VEV) of the Higgs field spontaneously breaks $\SUtwoLUoneY$ to $\UoneQ$. In the process, three of the four degrees of freedom associated with the complex Higgs doublet become the longitudinal polarizations of $\Wpm$ and $\Z$, giving them mass. The remaining degree of freedom is the physical Higgs boson. Expanding the Higgs field about its VEV produces Yukawa interactions such as those in Eq.~\ref{yukawa_terms}. These terms are interpreted as giving mass to the quarks and charged leptons mass, where the mass of each particle is given by the product of the Higgs VEV, $v$, and the coupling constant, $Y$.

    \noindent \begin{equation}
        \mathcal{L} = -\frac{Yv}{\sqrt{2}}\bar{\psi}\psi 
        \label{yukawa_terms}
    \end{equation}

@@

\section{Supersymmetry}
    \fxnote{seems like I should explicitly mention that naturalness favors lighter sparticles}  
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions\fxnote{I need to be better at explaining exactly what this means}. In its simplest form, SUSY results in one new boson for every SM fermion, one new fermion for every SM boson, and one new Higgs doublet.\fxnote{make sure I understand the new Higgs doublet} The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-$\frac{1}{2}$ superpartners to the SM bosons are given the name of their SM counterpart with ``ino" tacked on to the end (e.g. Higgsino or wino).\fxnote{describe charginos, neutralinos?} Collectively the superpartners to the SM particles are known as sparticles.\fxnote{cite martin}

@@

    \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [inline=(d.base), layered layout, medium, horizontal=b to c] {
        a -- [scalar] b,
        b -- [fermion, half left] c,
        c -- [fermion, half left] b,
        c -- [scalar] d,
    };
    \quad + \quad
    \feynmandiagram [baseline={(current bounding box.center)}, layered layout, medium, horizontal=b to c] {
        a -- [scalar] b,
        b -- [charged scalar, half left] c,
        c -- [charged scalar, half left] b,
        c -- [scalar] d,
    };
    \quad = \quad 0 \quad
    \caption{Caption!\fxnote{cite something}}
    \label{exact_susy_correction}
    \end{center} \end{figure}

    Exact SUSY also requires that sparticles have the same mass as their SM counterparts, so the lack of sparticles in collider searches implies that SUSY must be broken. With broken susy, the same diagram as above produces a correction of the form\fxnote{cite feng}

    \noindent \begin{equation}
        \delta \mHiggs^2 = \frac{3Y^2}{8\pi^2}\left(m_{\sfermion}^2 - m_{\f}^2 \right)\ln(\Lambda^2/m_{\sfermion}^2).
    \label{higgs_susy_correction}
    \end{equation}

@@

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs. Fig.~\ref{rp_and_rpv_diagrams} shows two possible processes, one where R-parity gaurantees a stable LSP and one where RPV produces a long-lived but stable LSP. A similar situation arises in GMSB models\fxnote{should I describe GMSB?} where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs.\fxnote{cite dimopoulos low energy}

@@

    \subsection{Electromagnetic Calorimeter}
        \fxnote{cite CMS TDR}
        \fxnote{I define EE \& EB but never use those definitions}
        After traversing the inner tracker, particles next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses \num{61200} lead tunstate crystals in the barrel and \num{7324} in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (\SI{80}{\percent} of light within \SI{25}{ns}), compact (radiation length = \SI{0.89}{cm}), fine-grained (Moli\`ere radius = \SI{2.2}{cm}), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (\SI{30}{photon\per\mega\electronvolt}), which necesitates photodetectors with intrinsic gain that work in magnetic fields.

@@

    \subsection{Hadronic Calorimeter}
        \fxnote{cite CMS TDR}
        \fxnote{define HB \& HF never to be used}
        The hadronic calorimeter (HCAL) uses \SI{3.7}{mm} plates of plastic scintillator interspersed within brass absorber to reconstruct the energy deposited during hadronic showers. Embedded wavelength-shifting fibers capture the scintillation light and transfer it to clear fibers to be read out by hybrid photodiodes.

@@

\section{LHC Searches for Long-lived Particles}
        \fxnote{mention which data was used for each search} 
    \subsection{Overview}
        \fxnote{steal jamie's cartoon?}  
        LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays, so LLP searches are often forced\fxnote{the searchers are forced, not the searches} to stray from the traditional analysis path in terms of object reconstruction\fxnote{really reconstruction? Maybe object definition?} and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that come to a stop before decaying. See Fig.~\ref{jamie_cartoon} for a cartoon of several possible LLP signatures\fxnote{probably a better way to word this sentence}.

@@

    \subsection{Displaced Vertices}
        \fxnote{cite CONF-2017-026}
        LLPs that decay before reaching the inner tracker can be identified with a displaced vertex (DV)\fxnote{DVs defined clearly enough in overview?}\fxnote{Is it still a DV if it decays in active material somewhere?}. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity.\fxnote{make sure I can explain why met and multiplicity} The search is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range\fxnote{Am I defining sensitivity range consistently across analyses?}.

@@

        After estimating $\num{0.02} \pm \num{0.02}$ background events, exactly zero are observed. The result is then interpreted as an upper limit on the gluino mass and pair-production cross section as a function of gluino lifetime, shown in Fig.~\ref{displaced_limit}. For an LSP neutralino mass of \SI{100}{\giga\electronvolt}, the gluino mass limit is approximately \SI{2000}{\giga\electronvolt} for lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range.

@@

    \subsection{Disappearing Tracks}
        \fxnote{cite CONF-2017-017}
        \fxnote{nice disappearing track cartoon}
        \fxnote{nice cartoon feynman diagrams of signal processes}
        Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos\fxnote{have these words been defined?}, such as many AMSB models\fxnote{why amsb? what is amsb?}, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate\fxnote{why? how typical?}) that eventaully decays to the neutralino LSP and, commonly, a pion\fxnote{why?}. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be reconstructed, which results in a disappearing track.

@@

        The authors consider two signal processes, both shown in Fig.\ref{disappearing_processes}: (1) electroweak gaugino production resulting in a disappearing track, an ISR jet, and MET and (2) gluino pair production resulting in a disappearing track, four jets, and MET. Process (1) could be the only gaugino production mode at the LHC if the gluino mass is kinematically unaccessable\fxnote{why?}, but process (2), which leads to chargino production in cascade decays, will be dominant if the gluino mass is within reach.

@@

    \subsection{Stopped Particles}
        \fxnote{Add calorimeter search?}  
        \fxnote{cite CMS PAS EXO-17-004}
        \fxnote{mention how stopped particles complement HSCP search?}
        \fxnote{too redundant with overview?}
        With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states using the full 2015 and 2016 datasets.

@@

    \subsection{Heavy Stable Charged Particles}
        \fxnote{cite CMS EXO-15-010}
        \fxnote{nice dE/dx plot (fig 1)}
        Particles that live for more than a couple ns\fxnote{how do I typeset standalone units?} \fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material\fxnote{redundant w/ overview. Probably better explained here than there}. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss, as shown in Fig.~\ref{hscp_dedx}.

@@

        HSCPs will often be misidentified or missed entirely by standard reconstruction algorithms and analysis criteria, which tend to assume $\beta \approx \num{1}$ and $\lvert q \rvert = e$. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models\fxnote{do I need figures for these?}: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with arbitrary electric charge. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY\fxnote{why?}.

@@

    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=0.8\textwidth]{images/atlas_susy_summary}
    \caption{Summary of current experimental limits on SUSY models from ATLAS searches.\fxnote{cite something}}
    \label{atlas_susy_summary}
    \end{center} \end{figure}
